cubble: An R Package for Structuring Spatio-temporal Data

### 2022-04-27 2-0-cube.Rmd
Comments and modifications 2-0-cube section

Spatio-temporal data can be conceptualised using a cubical data model with three axes, typically time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation: @lu_multidimensional_2018 shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; @bach_review_2014 reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but refers to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisation still applies to our spatio-temporal data manipulation and visualisation.

The most common data cube using the three axes (time, latitude, longitude)  could be considered as taking snapshots of the space and stacking on time. Here we will define the spatio-temporal cube slightly differently with the three axes (time, site, variable), as illustrated in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array,  may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) The `mutate()` syntax for creating new variables is no longer valid since the operation can no longer define how a created 1D vector should be added to the array.
\textcolor{blue}{Ursula: this seems too specific, you should be able to make the same argument without referencing the mutate function?}

>>>>

Spatio-temporal data can be conceptualised using a cubical data model with three axes, typically time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation purposes: @lu_multidimensional_2018 shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; @bach_review_2014 reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but refers to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisation still applies to our spatio-temporal data manipulation and visualisation.

The most common data cube using the three axes (time, latitude, longitude)  could be considered as taking snapshots of the space and stacking on time. Here we will define the spatio-temporal cube slightly differently with the three axes (time, site, variable), as illustrated in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the workhorse behind \pkg{cubble}.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array,  may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) The `mutate()` syntax for creating new variables is no longer valid since the operation can no longer define how a created 1D vector should be added to the array.
\textcolor{blue}{Ursula: this seems too specific, you should be able to make the same argument without referencing the mutate function?}
\textcolor{green}{How about the other functions from dplyr?, the same principle would apply right?}

### 2022-04-27 arxiv/cubble-arxiv.tex
Comments and modifications 2-0-cube section

Spatio-temporal data can be conceptualised using a cubical data model with three axes, typically time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation: \protect\hyperlink{ref-lu_multidimensional_2018}{Lu, Appel, and Pebesma} (\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; \protect\hyperlink{ref-bach_review_2014}{Bach et al.} (\protect\hyperlink{ref-bach_review_2014}{2014}) reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data,'' but refers to ``an abstract 2D substrate that is used to visualize data at a specific time.'' Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisation still applies to our spatio-temporal data manipulation and visualisation.

The most common data cube using the three axes (time, latitude, longitude) could be considered as taking snapshots of the space and stacking on time. Here we will define the spatio-temporal cube slightly differently with the three axes (time, site, variable), as illustrated in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array, may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) The \texttt{mutate()} syntax for creating new variables is no longer valid since the operation can no longer define how a created 1D vector should be added to the array.
\textcolor{blue}{Ursula: this seems too specific, you should be able to make the same argument without referencing the mutate function?}

>>>>

Spatio-temporal data can be conceptualised using a cubical data model with three axes, typically time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation purposes: \protect\hyperlink{ref-lu_multidimensional_2018}{Lu, Appel, and Pebesma} (\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; \protect\hyperlink{ref-bach_review_2014}{Bach et al.} (\protect\hyperlink{ref-bach_review_2014}{2014}) reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data,'' but refers to ``an abstract 2D substrate that is used to visualize data at a specific time.'' Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisation still applies to our spatio-temporal data manipulation and visualisation.

The most common data cube using the three axes (time, latitude, longitude) could be considered as taking snapshots of the space and stacking on time. Here we will define the spatio-temporal cube slightly differently with the three axes (time, site, variable), as illustrated in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the workhorse behind \pkg{cubble}.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array, may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) The \texttt{mutate()} syntax for creating new variables is no longer valid since the operation can no longer define how a created 1D vector should be added to the array.
\textcolor{blue}{Ursula: this seems too specific, you should be able to make the same argument without referencing the mutate function?}
\textcolor{green}{How about the other functions from dplyr?, the same principle would apply right?}
